Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-06-04 paper/AlignedIsotropicGW.tex
O2 --> GW170104 where appropriate.

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} \textbf{Approximate posteriors on
      $\chieff$ from the Advanced LIGO O1 and O2
      observations\cite{O1-BBH,PhysRevLett.118.221101}.}  We
    approximate the posteriors reported using Gaussians with the same
    median and 90\% credible interval.  It is notable that none of the
    $\chieff$ posteriors support high \ac{BH} spin magnitudes with
    aligned spins, suggested by observations of stellar-mass black
    holes in X-ray binaries\cite{2015PhR...548....1M}.}
\end{figure}

@@

We fit hierarchical models of the four existing LIGO O1 and O2
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the spin magnitude
distribution is fixed but the angular distribution is a weighted sum
of the isotropic and aligned distribution.  The evidence, or marginal
likelihood, for each of the models is shown in Figure
\ref{fig:O1-odds}.  For all three magnitude distributions, the mixture
models' posterior on the mixing fraction peaks at 100\% isotropic,
which explains why the zero-parameter, pure-isotropic models are
preferred over the single-parameter mixture models for every magnitude
distribution with this data set.  Not surprisingly, given the small
$\chieff$ values in the three detected systems, the most-favoured
model among those with an isotropic angular distribution has the
``low'' magnitude distribution; the most favoured model among those
with an aligned distribution also has the ``low'' magnitude
distribution.  The odds ratio between the ``low'' aligned and ``low''
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
binary black hole systems having zero spin is ruled out by the
GW151226 measurements, which bound at least one black hole to have
spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

\begin{figure}
  \plotone{../plots/Wills_evidence_ratio_figure_with_mixture_models}
  \caption{\textbf{Odds ratios among our models using the
      approximations to the posteriors on $\chieff$ from the O1 and O2
      observations shown in Figure \ref{fig:O1-posteriors}.}  The flat
    (``F''), high (``H''), and low (``L'') spin magnitude
    distributions (see Eq.\ \eqref{eq:magnitude-dists}) are paired
    with isotropic (``I'') and aligned (``A'') angular distributions,
    as well as a mixture model of the two (``$\mathrm{M}$'').  The
    most-favoured models have the ``low'' distribution of spin
    magnitudes.  The odds ratio between the best aligned and best
    isotropic models is $\OOneOddsIsoAligned$, or
    $\OOneSigmaIsoAligned\sigma$.  For all magnitude distributions the
    pure-isotropic models are preferred over the mixture models;
    correspondingly, the posterior on the mixture fraction peaks at
    100\% isotropic.}
  \label{fig:O1-odds}
\end{figure}

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
for the spin distribution in addition to the four existing detections
from O1 and O2, with observational uncertainties drawn randomly from
the three Gaussian widths used to approximate the $\chieff$ posteriors
in Figure \ref{fig:O1-posteriors}\footnote{The measurement uncertainty
  in $\chieff$ depends on the other parameters of the merging binary
  black hole system, particularly on the mass ratio.  Our assumption
  about future observational uncertainties is appropriate if the
  parameters of the three detected events are representative of the
  parameters of future detections. See Methods Section
  \ref{sec:chi-eff-precision} for further discussion.}, we find the
odds ratios shown in Figure \ref{fig:O2-predictions}.  We find that
most scenarios with an additional 10 detections allow the simulated
angular distribution to be inferred with greater than $5\sigma$
($2.9 \times 10^{-7}$ odds) credibility; and in the most pessimistic
case the distinction is typically $\OTwoSigmaIsoAlignedMin\sigma$
($\OTwoOddsIsoAlignedMin$ odds ratio).  While such future detections
should permit a confident distinction between \emph{angular}
distributions, we would remain much less certain about the
\emph{magnitude} distribution among the three options considered here
until we have a larger number of
observations\cite{2017arXiv170306869F,2017arXiv170306223G}.  In Figure
\ref{fig:O2-predictions}, the odds ratio between different magnitude
distributions with the same angular distribution is much closer to
unity than the odds ratio between angular distributions.

\begin{figure}
  \plotone{../plots/six-way-O2-model-selection}
  \caption{\label{fig:O2-predictions} \textbf{Distribution of odds
      ratios predicted with 10 additional observations above the four
      discussed above.}  Each panel corresponds to additional
    observations drawn from one of the $\chieff$ distribution models.
    The model from which the additional observations are drawn is
    outlined in red.  The height of the blue bar gives the median odds
    ratio relative to the model from which the additional observations
    are drawn; the green line gives the 68\% ($1 \sigma$) symmetric
    interval of odds ratios over 1000 separate draws from the model
    distribution.  The closest median ratio between the most-favoured
    isotropic model and the most-favoured aligned model is
    $\OTwoOddsIsoAlignedMin$, corresponding to
    $\OTwoSigmaIsoAlignedMin\sigma$ preference for the correct angular
    distribution; most models result in more than $5\sigma$ preference
    for the correct angular distribution.  Because the four existing
    observations are included in each data set the ``correct'' model
    is not necessarily preferred over the others, particularly when
    that model uses the ``high'' magnitude distribution, which is
    strongly dis-favoured from the O1 and O2 observations alone.}
\end{figure}

@@

%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../plots/posterior_on_isotropic_fraction_with_GW170104.png}
\caption{\textbf{Fraction of the BBH population coming from an
    isotropic distribution under a mixture model.} The dotted line
  shows the flat prior on the fraction of BBHs coming from an
  isotropic distribution, $f_i$, under the mixture model. The 3 red
  lines show the posterior on $f_i$ after O1 and O2 with our various
  assumptions regarding BH spin magnitudes.  The solid line shows the
  posterior assuming that all BHs have their spin magnitude drawn from
  the ``flat'' distribution. The dashed line assumes the ``high'' BH
  spin magnitude distribution $p(a) = 2a$. The dot-dash line assumes
  the ``low'' distribution $p(a) = 2(1-a)$.  We see that for a wide
  range of assumptions regarding BH spin magnitudes, the fraction
  coming from an isotropic distribution $f_i$ peaks at 1.}
\label{fig:mixture_fraction_posterior}
\end{figure}
%

@@

We fit a mixture model\cite{Stevenson:2017spin} (labelled model
'M' in Figure~\ref{fig:O1-odds}) where a fraction $f_i$ of BBHs have
spins drawn from an isotropic distribution, whilst a fraction
$1 - f_i$ have their spins aligned with the orbital angular
momentum. We assume a flat prior on the fraction $f_i$. To test the
robustness of our result, we vary the distribution we assume for BH
spin magnitude distributions as with the aligned and isotropic
models. We use the ``flat'', ``high'' and ``low'' distributions
(Equation~\ref{eq:magnitude-dists}), assuming all BHs have their spin
magnitude drawn from the same distribution for both the aligned and
isotropic populations. We calculate and plot the posterior on $f_i$
given by Equation~\ref{eq:hierarchical-posterior} ($f_i=\lambda$ in
the derivation) in Extended Data
Figure~\ref{fig:mixture_fraction_posterior}. We find the mean fraction
of BBHs coming from an isotropic distribution is 0.70, 0.77 and 0.81
assuming the ``low'', ``flat'' and ``high'' distributions for spin
magnitudes respectively, compared to the prior mean of 0.5. The lower
90\% limits are 0.38, 0.51 and 0.60 respectively, compared to the
prior of 0.1. In all cases, the posterior peaks at $f_i = 1$. Thus,
for these spin magnitude distributions we find that the current O1 and
O2 LIGO observations constrain the majority of BBHs to have their
spins drawn from an isotropic distribution. The evidence ratios of these mixture models to the isotropic distribution with ``low'' spin magnitudes are 0.43, 0.20 and 0.10 for the ``low'', ``flat''
and ``high'' spin magnitude models. Thus we cannot rule out a mixture with
the current data.  If several different components contribute
significantly to the true spin distribution it may take tens to
hundreds of detections to accurately determined the mixing fraction,
depending on the distribution of spin
magnitudes\cite{2017CQGra..34cLT01V,Stevenson:2017spin}.

>>>>

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} \textbf{Approximate posteriors on
      $\chieff$ from the Advanced LIGO O1 and GW170104
      observations\cite{O1-BBH,PhysRevLett.118.221101}.}  We
    approximate the posteriors reported using Gaussians with the same
    median and 90\% credible interval.  It is notable that none of the
    $\chieff$ posteriors support high \ac{BH} spin magnitudes with
    aligned spins, suggested by observations of stellar-mass black
    holes in X-ray binaries\cite{2015PhR...548....1M}.}
\end{figure}

@@

We fit hierarchical models of the four existing LIGO O1 and GW170104
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the spin magnitude
distribution is fixed but the angular distribution is a weighted sum
of the isotropic and aligned distribution.  The evidence, or marginal
likelihood, for each of the models is shown in Figure
\ref{fig:O1-odds}.  For all three magnitude distributions, the mixture
models' posterior on the mixing fraction peaks at 100\% isotropic,
which explains why the zero-parameter, pure-isotropic models are
preferred over the single-parameter mixture models for every magnitude
distribution with this data set.  Not surprisingly, given the small
$\chieff$ values in the three detected systems, the most-favoured
model among those with an isotropic angular distribution has the
``low'' magnitude distribution; the most favoured model among those
with an aligned distribution also has the ``low'' magnitude
distribution.  The odds ratio between the ``low'' aligned and ``low''
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
binary black hole systems having zero spin is ruled out by the
GW151226 measurements, which bound at least one black hole to have
spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

\begin{figure}
  \plotone{../plots/Wills_evidence_ratio_figure_with_mixture_models}
  \caption{\textbf{Odds ratios among our models using the
      approximations to the posteriors on $\chieff$ from the O1 and GW170104
      observations shown in Figure \ref{fig:O1-posteriors}.}  The flat
    (``F''), high (``H''), and low (``L'') spin magnitude
    distributions (see Eq.\ \eqref{eq:magnitude-dists}) are paired
    with isotropic (``I'') and aligned (``A'') angular distributions,
    as well as a mixture model of the two (``$\mathrm{M}$'').  The
    most-favoured models have the ``low'' distribution of spin
    magnitudes.  The odds ratio between the best aligned and best
    isotropic models is $\OOneOddsIsoAligned$, or
    $\OOneSigmaIsoAligned\sigma$.  For all magnitude distributions the
    pure-isotropic models are preferred over the mixture models;
    correspondingly, the posterior on the mixture fraction peaks at
    100\% isotropic.}
  \label{fig:O1-odds}
\end{figure}

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
for the spin distribution in addition to the four existing detections
from O1 and GW170104, with observational uncertainties drawn randomly
from the three Gaussian widths used to approximate the $\chieff$
posteriors in Figure \ref{fig:O1-posteriors}\footnote{The measurement
  uncertainty in $\chieff$ depends on the other parameters of the
  merging binary black hole system, particularly on the mass ratio.
  Our assumption about future observational uncertainties is
  appropriate if the parameters of the three detected events are
  representative of the parameters of future detections. See Methods
  Section \ref{sec:chi-eff-precision} for further discussion.}, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
simulated angular distribution to be inferred with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we would remain much less
certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of
observations\cite{2017arXiv170306869F,2017arXiv170306223G}.  In Figure
\ref{fig:O2-predictions}, the odds ratio between different magnitude
distributions with the same angular distribution is much closer to
unity than the odds ratio between angular distributions.

\begin{figure}
  \plotone{../plots/six-way-O2-model-selection}
  \caption{\label{fig:O2-predictions} \textbf{Distribution of odds
      ratios predicted with 10 additional observations above the four
      discussed above.}  Each panel corresponds to additional
    observations drawn from one of the $\chieff$ distribution models.
    The model from which the additional observations are drawn is
    outlined in red.  The height of the blue bar gives the median odds
    ratio relative to the model from which the additional observations
    are drawn; the green line gives the 68\% ($1 \sigma$) symmetric
    interval of odds ratios over 1000 separate draws from the model
    distribution.  The closest median ratio between the most-favoured
    isotropic model and the most-favoured aligned model is
    $\OTwoOddsIsoAlignedMin$, corresponding to
    $\OTwoSigmaIsoAlignedMin\sigma$ preference for the correct angular
    distribution; most models result in more than $5\sigma$ preference
    for the correct angular distribution.  Because the four existing
    observations are included in each data set the ``correct'' model
    is not necessarily preferred over the others, particularly when
    that model uses the ``high'' magnitude distribution, which is
    strongly dis-favoured from the O1 and GW170104 observations
    alone.}
\end{figure}

@@

%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../plots/posterior_on_isotropic_fraction_with_GW170104.png}
\caption{\textbf{Fraction of the BBH population coming from an
    isotropic distribution under a mixture model.} The dotted line
  shows the flat prior on the fraction of BBHs coming from an
  isotropic distribution, $f_i$, under the mixture model. The 3 red
  lines show the posterior on $f_i$ after O1 and GW170104 with our
  various assumptions regarding BH spin magnitudes.  The solid line
  shows the posterior assuming that all BHs have their spin magnitude
  drawn from the ``flat'' distribution. The dashed line assumes the
  ``high'' BH spin magnitude distribution $p(a) = 2a$. The dot-dash
  line assumes the ``low'' distribution $p(a) = 2(1-a)$.  We see that
  for a wide range of assumptions regarding BH spin magnitudes, the
  fraction coming from an isotropic distribution $f_i$ peaks at 1.}
\label{fig:mixture_fraction_posterior}
\end{figure}
%

@@

We fit a mixture model\cite{Stevenson:2017spin} (labelled model 'M' in
Figure~\ref{fig:O1-odds}) where a fraction $f_i$ of BBHs have spins
drawn from an isotropic distribution, whilst a fraction $1 - f_i$ have
their spins aligned with the orbital angular momentum. We assume a
flat prior on the fraction $f_i$. To test the robustness of our
result, we vary the distribution we assume for BH spin magnitude
distributions as with the aligned and isotropic models. We use the
``flat'', ``high'' and ``low'' distributions
(Equation~\ref{eq:magnitude-dists}), assuming all BHs have their spin
magnitude drawn from the same distribution for both the aligned and
isotropic populations. We calculate and plot the posterior on $f_i$
given by Equation~\ref{eq:hierarchical-posterior} ($f_i=\lambda$ in
the derivation) in Extended Data
Figure~\ref{fig:mixture_fraction_posterior}. We find the mean fraction
of BBHs coming from an isotropic distribution is 0.70, 0.77 and 0.81
assuming the ``low'', ``flat'' and ``high'' distributions for spin
magnitudes respectively, compared to the prior mean of 0.5. The lower
90\% limits are 0.38, 0.51 and 0.60 respectively, compared to the
prior of 0.1. In all cases, the posterior peaks at $f_i = 1$. Thus,
for these spin magnitude distributions we find that the current O1 and
GW170104 LIGO observations constrain the majority of BBHs to have
their spins drawn from an isotropic distribution. The evidence ratios
of these mixture models to the isotropic distribution with ``low''
spin magnitudes are 0.43, 0.20 and 0.10 for the ``low'', ``flat'' and
``high'' spin magnitude models. Thus we cannot rule out a mixture with
the current data.  If several different components contribute
significantly to the true spin distribution it may take tens to
hundreds of detections to accurately determined the mixing fraction,
depending on the distribution of spin
magnitudes\cite{2017CQGra..34cLT01V,Stevenson:2017spin}.
